---
layout: post
title: "Dungeons and Dragons, Stats for 6/18/2015"
date:  2015-06-21 22:46:43
comments: true
categories: R, ggplot2, DnD
published: true
author: dyg
---

So me and my friends play Dungeons and Dragons every week. A few weeks in we 
decided it would be a good idea to start recording our d20 rolls. This came from
one of my friends being declared as highly bimodel in his rolls, i.e. he rolls 
close to a 1 or a 20. I'm no fan of confirmation bias, and having a love for
stats, R and data graphics I thought this would be a fun analysis. I will most
likely post these types of analyses every week or so depending on how busy I am.

##A look at the dataset
```{r dataLoad, echo = TRUE}
#all the states are kept in a google docs sheet, which can be downloaded as a 
#csv
df <- read.csv("Roll_Stats_06212015.csv", header = T, sep = ",")

#first look
str(df)
```

Ah, a few considerations before we begin plotting away. First the Person 
variable needs to be cleaned up, I know for a fact that we have 4 players, but 
R is listing 10 uniques. To me this means there are probably some typos or
leading and trailing spaces. Secondly, the Date variable isn't a date. This 
should do it:

```{r cleanData, echo = TRUE, message=FALSE}
#load some libs
library(stringr)
library(dplyr)

#Now do the cleaning
df$Person <- df$Person %>%
  tolower() %>%
    str_trim() %>%
      as.factor()
  
df$Date <- as.Date(as.character(df$Date), format = "%m/%d/%Y")

#have a quick look at the cleaned dataset
str(df)
```  

Perfect. First I want see how many total rolls have happened, and how many weeks
we have been recording data:

```{r, basicStats, echo = TRUE}
#Observe that each row is a roll
head(df)

#now what are the dimensions of the dataset
length(df$Roll_Value)
```

Now how many rolls has everyone made to date, by date, and what were they, 
specifically?

```{r, rollByDate, echo = TRUE}
#Total rolls by Person
table(df$Person)

#Rolls by person/date
table(df$Date, df$Person)

#Roll count by Person and Roll
table(df$Roll_Value, df$Person)
```

Thats nice but its really just for someone who wants detail that a plot isn't 
going to provide. A few things of note, me (Jeff) have rolled the second most
one's after mike. I would expect this data to be affected by play style, class,
and most importantly luck. A quick note, the player "mel" is a halfling who has
the feat "lucky". Basically, whenever she rolls a nat 1 she can re-roll. Being
the data scribe in the group my practice has been to record both d20 rolls.

##Pretty Graphics
Ok now on to the fun stuff. I want to see each players total distributions, and
how each players roll behavior has changed each week.

```{r playerDistributions, echo = TRUE, fig.height=8, fig.width=11}
#load ggplot
library(ggplot2)

#Make histrograms faceted by person
ggplot(data = df, aes(x = Roll_Value, fill = Person)) +
  geom_histogram(stat = "bin") +
  facet_grid(Person ~ .) +
  ggtitle("Rolls by Player") +
  scale_x_discrete(breaks = seq(1, 20)) 
```

Now lets see how each players rolls have changed over time. For this I'm going
with a tried and true time series plot, plotting the mean and median roll
for each player distribution.

```{r playerChanges, echo = TRUE, fig.height = 8, fig.width = 11}
#load dplyr
library(dplyr)

#do some statistics
simpleStat <- function(fun, geom="point", ...) {
  stat_summary(fun.y = fun, geom=geom, ...)
}

myQuantile <- function(x) {
  out <- quantile(x, probs = c(0.25, 0.75))
  out
}

myMode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

#now generate plot
ggplot(df, aes(x = factor(Date), y = Roll_Value, fill = Person)) +
  xlab("Date") +
  geom_violin(trim = FALSE, adjust = 0.5) +
  simpleStat(myMode, size = 3) +
  simpleStat(mean, size = 1.5) +
  facet_grid(Person ~ .) +
  theme(legend.position = "none") +
  ggtitle("Player Roll Distributions by Date")
```

The plot is a violin plot for each week of data given (for more on Violin Plots
go [here][1]), the large dot in each violin plot is the mode of that players
rolls for that week, the smaller dot is the mean of that players rolls for that 
week. Subjectively, mode seems to be a better indicator of how well I percieve 
I roll in a given week. For instance, I remember think "I rolled like complete 
crap today", two weeks ago on 6/12. The mode displays this more clearly than 
the mean.


Lastly I want a line plot to literally connect the dots over each week for each
player:

```{r finalPlot, echo = TRUE, fig.height = 8, fig.width = 11}
#generate summary dataset
df.change <- df %>%
  group_by(Date, Person) %>%
    summarise(personMean = mean(Roll_Value, na.rm = TRUE),
              personMode = myMode(Roll_Value))

#check the new dataset
head(df.change)

#great now lets melt it with tidyr
library(tidyr)
df.tidy <- df.change %>%
  gather(stat, value, -Person, -Date)

#again, a quick look
head(df.tidy)

#plot time series
ggplot(df.tidy, aes(x = factor(Date), y = value, group = stat, 
                    color = stat)) +
  geom_line(aes(linetype = stat), size = 1.5) +
  facet_grid(Person ~ .) +
  ggtitle("Mean and Mode by Player, by Date") +
  ylab("Roll Value") + 
  xlab("Date")
```

##Conclusions
My friend Will thinks he encounters low rolls every three weeks. Well that is
objectively not true according to this dataset. However Will is the most 
consistent player. Looking at his mean and mode he shows the least variation of
anyone. One interesting detail is that
Mike is decidedly bimodal, is mode oscillates almost perfectly on a two week
basis. Not only this but he has rolled the most 1's, and the second most 20's. 